---
title: "Survival Analysis of Colon Cancer Adjuvant Therapy Trial"
author: "Van-Liêm PHAM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.path = here::here("figures", ""), fig.width = 8, fig.height = 6)
```

```{r packages}
library(survival)
library(survminer)
library(dplyr)
library(ggplot2)
library (here)
```

```{r data}
data(colon)

# etype == 2 : événement de décès (etype == 1 = récidive).
# Sans ce filtre, chaque patient apparaît deux fois — biais majeur.
df <- colon %>%
  filter(etype == 2) %>%
  mutate(
    rx       = factor(rx, levels = c("Obs", "Lev", "Lev+5FU")),
    sex      = factor(sex, labels = c("Female", "Male")),
    obstruct = factor(obstruct, labels = c("No", "Yes")),
    perfor   = factor(perfor,   labels = c("No", "Yes")),
    adhere   = factor(adhere,   labels = c("No", "Yes")),
    differ   = factor(differ,   labels = c("Well", "Moderate", "Poor")),
    extent   = factor(extent,   labels = c("Submucosa", "Muscle", "Serosa", "Adjacent")),
    node4    = factor(node4,    labels = c("<=4 nodes", ">4 nodes")),
    surg     = factor(surg,     labels = c("Short", "Long"))
  )

dim(df)
summary(df$rx)
```

```{r descriptive}
# Distribution du temps de suivi
ggplot(df, aes(x = time / 365.25)) +
  geom_histogram(bins = 30, fill = "#1F3A5F", colour = "white") +
  labs(x = "Follow-up (years)", y = "Number of patients",
       title = "Distribution of follow-up time") +
  theme_minimal()

# Taux d'événement par bras de traitement
df %>%
  group_by(rx) %>%
  summarise(n = n(),
            deaths = sum(status),
            event_rate = round(mean(status), 3))
```

```{r km_overall}
fit_overall <- survfit(Surv(time, status) ~ 1, data = df)
ggsurvplot(fit_overall, data = df,
           conf.int = TRUE,
           xlab = "Time (days)", ylab = "Overall survival probability",
           palette = "#1F3A5F",
           ggtheme = theme_minimal())
```

```{r km_treatment}
fit_treatment <- survfit(Surv(time, status) ~ rx, data = df)

ggsurvplot(fit_treatment, data = df,
           conf.int = TRUE,
           pval = TRUE,           # p-value du test du log-rank
           risk.table = TRUE,     # tableau du nombre de sujets à risque
           xlab = "Time (days)", ylab = "Overall survival probability",
           legend.title = "Treatment",
           palette = c("#888888", "#3498DB", "#1F3A5F"),
           ggtheme = theme_minimal())
```

```{r logrank_pairwise}
# Le test global ne dit pas quel bras porte le bénéfice.
# Contrastes par paire contre le bras observation.
pairwise_logrank <- function(arm) {
  sub <- df %>%
    filter(rx %in% c("Obs", arm)) %>%
    mutate(rx = droplevels(rx))
  lr <- survdiff(Surv(time, status) ~ rx, data = sub)
  p  <- 1 - pchisq(lr$chisq, df = 1)
  data.frame(comparison = paste(arm, "vs Obs"),
             chisq = round(lr$chisq, 2),
             p = format.pval(p, digits = 3, eps = 1e-4))
}

bind_rows(lapply(c("Lev", "Lev+5FU"), pairwise_logrank))
```

```{r cox_univariate}
cox_uni <- coxph(Surv(time, status) ~ rx, data = df)
summary(cox_uni)
```

```{r cox_multivariate}
cox_multi <- coxph(Surv(time, status) ~ rx + age + sex + obstruct +
                                          perfor + adhere + differ +
                                          extent + node4 + surg,
                   data = df)
summary(cox_multi)
```

```{r missing_data_check}
cat("Univariate model   — n =", cox_uni$n,   ", events =", cox_uni$nevent, "\n")
cat("Multivariate model — n =", cox_multi$n, ", events =", cox_multi$nevent, "\n")
cat("Rows dropped due to missing covariates:", nrow(df) - cox_multi$n, "\n\n")

# Laisser le code nommer la variable responsable plutôt que de l'affirmer.
covars    <- all.vars(formula(cox_multi))[-(1:2)]
na_counts <- colSums(is.na(df[, covars, drop = FALSE]))
print(na_counts[na_counts > 0])
```

```{r sensitivity_missing}
cox_nodiffer <- coxph(Surv(time, status) ~ rx + age + sex + obstruct +
                        perfor + adhere + extent + node4 + surg,
                      data = df)

data.frame(
  model = c("With differ (complete cases)", "Without differ (all patients)"),
  n     = c(cox_multi$n, cox_nodiffer$n),
  HR    = round(c(summary(cox_multi)$conf.int["rxLev+5FU", "exp(coef)"],
                  summary(cox_nodiffer)$conf.int["rxLev+5FU", "exp(coef)"]), 3)
)
```

```{r forest_plot}
ggforest(cox_multi, data = df)
```

```{r ph_test}
ph_test <- cox.zph(cox_multi)
ph_test

ph_global_p <- ph_test$table["GLOBAL", "p"]
```

```{r ph_plot, fig.height=10}
ggcoxzph(ph_test)
```

# Interpretation

After adjusting for established prognostic factors (age, sex, bowel obstruction,
perforation, tumour differentiation, extent of invasion, nodal involvement, and
time from surgery; n = `r cox_multi$n`, events = `r cox_multi$nevent`), the
levamisole + 5-FU combination is associated with a hazard ratio of
`r round(summary(cox_multi)$conf.int["rxLev+5FU", "exp(coef)"], 2)`
(95% CI `r round(summary(cox_multi)$conf.int["rxLev+5FU", "lower .95"], 2)`–`r round(summary(cox_multi)$conf.int["rxLev+5FU", "upper .95"], 2)`,
p = `r format.pval(summary(cox_multi)$coefficients["rxLev+5FU", "Pr(>|z|)"], digits = 3, eps = 1e-4)`)
compared to observation alone — a `r round(100 * (1 - summary(cox_multi)$conf.int["rxLev+5FU", "exp(coef)"]))`% reduction in the hazard of death,
against the 33% reduction in overall death rate reported for stage C patients
in the original trial (Moertel et al., 1990).

`r if (ph_global_p >= 0.05) sprintf("The proportional hazards assumption was not rejected (global test p = %s).", format.pval(ph_global_p, digits = 3, eps = 1e-4)) else sprintf("The proportional hazards assumption was rejected globally (p = %s); see the cox.zph output above for the covariates concerned.", format.pval(ph_global_p, digits = 3, eps = 1e-4))`

```{r session}
sessionInfo()
```
